A prototype software framework for transparent, reusable and updatable computational health economic models

### 2023-10-22 Reports/Manuscript/Submission.tex
problems with table and figure placement

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{float}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}
\usepackage{lscape}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\usepackage[none]{hyphenat}
\usepackage[export]{adjustbox}
\usepackage{multirow}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={A prototype software framework for transparent, reusable and updatable computational health economic models},
  pdfauthor={Matthew P Hamilton1,2,; Caroline X Gao2,3,1; Glen Wiesner4; Kate M Filia2,3; Jana M Menssink2,3; Petra Plencnerova5; David Baker2,3; Patrick D McGorry2,3; Alexandra Parker6,3; Jonathan Karnon7; Sue M Cotton2,3; Cathrine Mihalopoulos1},
  pdfkeywords={computational models, ethics of modelling, health economics, mental disorders, open-source models, software frameworks},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{A prototype software framework for transparent, reusable and updatable computational health economic models}
\author{Matthew P Hamilton\textsuperscript{1,2,*} \and Caroline X Gao\textsuperscript{2,3,1} \and Glen Wiesner\textsuperscript{4} \and Kate M Filia\textsuperscript{2,3} \and Jana M Menssink\textsuperscript{2,3} \and Petra Plencnerova\textsuperscript{5} \and David Baker\textsuperscript{2,3} \and Patrick D McGorry\textsuperscript{2,3} \and Alexandra Parker\textsuperscript{6,3} \and Jonathan Karnon\textsuperscript{7} \and Sue M Cotton\textsuperscript{2,3} \and Cathrine Mihalopoulos\textsuperscript{1}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
\textbf{Summary: } Most health economic analyses are undertaken with the aid of computers. However, the ethical dimensions of implementing health economic models as software (or computational health economic models (CHEMs)) are poorly understood. We propose that developers and funders of CHEMs share ethical responsibilities to (i) establish socially acceptable user requirements and design specifications; (ii) ensure fitness for purpose; and (iii) support socially beneficial use. We further propose that a transparent (T), reusable (R) and updatable (U) CHEM is suggestive of a project team that has largely fulfilled these responsibilities. We propose six criteria for assessing CHEMs: (T1) software files are open access; (T2) project team contributions and judgments are easily identified; (R1) programming practices promote generalisability and transferability; (R2) licenses restrict only unethical reuse; (U1) maintenance infrastructure is in place; and (U2) new releases are systematically retested and appropriately deprecated. To facilitate CHEMs that meet TRU criteria, we have developed a prototype software framework in the open-source programming language R. The framework comprises six code libraries for authoring CHEMs, supplying CHEMs with data and undertaking analyses with CHEMs. The prototype software framework integrates with services for software development and research data archiving. We determine that an initial set of youth mental health CHEMs we developed with the prototype software framework wholly meet criteria T1-2, R1-2 and U1 and partially meet criterion U2. Our assessment criteria and prototype software framework can help inform and improve ethical implementation of CHEMs. Resource barriers to ethical CHEM practice should be addressed by research funders. \newline \newline \textbf{Code: } Visit \url{https://www.ready4-dev.com} for more information about how to find, install and apply the prototype software framework and CHEMs developed with it. \newline \newline
\end{abstract}

\textsuperscript{1} School of Public Health and Preventive Medicine, Monash University, Clayton, Australia\\
\textsuperscript{2} Orygen, Parkville, Australia\\
\textsuperscript{3} Centre for Youth Mental Health; The University of Melbourne, Parkville, Australia\\
\textsuperscript{4} Heart Foundation, Melbourne, Australia\\
\textsuperscript{5} headspace National Youth Mental Health Foundation, Melbourne, Australia\\
\textsuperscript{6} Institute for Health and Sport, Victoria University, Footscray, Australia\\
\textsuperscript{7} Flinders University, Adelaide, Australia

\textsuperscript{*} Correspondence: \href{mailto:matthew.hamilton1@monash.edu}{Matthew P Hamilton \textless{}\href{mailto:matthew.hamilton1@monash.edu}{\nolinkurl{matthew.hamilton1@monash.edu}}\textgreater{}}

@@

Health economics is a discipline concerned with problems that arise due to scarce resources, such as how to value health and healthcare, allocate healthcare budgets and configure health services {[}1{]}. In seeking to solve these problems, health economists typically use models which are simplified and selective representations of systems that are believed to influence human health. A health economic model should be capable of being described using words and figures (a conceptual model), equations (a mathematical model) and software (a computational model). Health economic scientific manuscripts typically describe a model using its conceptual and mathematical representations but report results that have been generated by its computational representation (i.e., the format that allows computers to perform calculations). The conceptual, mathematical and computational representations of a model are assumed to be isomorphic. However, the ease of independently assessing the validity of this assumption depends in part on how a health economic model has been implemented computationally.

The set of software files implementing a health economic model's structure, data and algorithms can be called a computational health economic model (CHEM). CHEMs can be developed using commercial modelling software or open-source programming languages such as Python {[}2{]} or R {[}3{]}. An advantage of commercial modelling software is that users often require limited or no programming skills to develop and apply CHEMs. However, open-source software programming languages may facilitate the development of CHEMs that are more transparent, reusable and updatable {[}4,5{]}. The decision about what type of software development platform to use must balance ethical considerations and resource constraints (see Table \ref{tab:proscons}).

\addlinespace[0.3em]
\multicolumn{3}{c}{\textbf{Updatability}}\\
\hspace{1em}\multirow{-7}{7em}{\raggedright\arraybackslash \textbf{Ethical}} & (-) Selective modification with appropriate attribution can be difficult. & (-) Facilitates collaborative development and maintenance.\\
\bottomrule
\end{tabular}
\end{table}

Open-source CHEMs remain rare {[}6,7{]}. The software used to develop CHEMs was identified as a barrier to open-source model implementations by participants in a 2020 survey of health economists {[}5{]}. Other reported barriers relate to difficulties in generalising models, updating models and transferring model data, concerns about the level of public access and lack of interest from decision-makers, and legal, confidentiality and security considerations {[}5{]}.

Better software frameworks could potentially help address some of the barriers to open-source CHEM development. A software framework is a shared common technology used by developers to collaboratively author software and which is not typically visible to software end-users {[}8{]}. A software framework provides a foundation for developing multiple software applications with shared resources (e.g., code and data files), that can be modified to suit specific needs. Advantages of using software frameworks include facilitating code reuse and extension, promoting good programming practice and the capability to provide enhanced functionality and performance without additional effort by developers {[}9{]}. Software frameworks have widely been developed and implemented in data science, for example, PyTorch {[}10{]} for machine learning using Python.

A high-level software framework for implementing open-source CHEMs in R has previously been developed with the primary aim of improving model transparency {[}11{]}. However, to gain widespread adoption, a software framework for CHEMs may need to provide tools to help model developers meet a broader range of ethical practice requirements. There are multiple challenges to overcome. Firstly, the ethical responsibilities of public health computational modellers remain poorly understood {[}12{]} and the computational modelling landscape is changing fast {[}13{]}. Secondly, software frameworks can be challenging and time consuming to create and may become excessively complex over time {[}9{]}. Thirdly, software frameworks can be difficult to learn, often requiring model developers to undergo specialist training {[}8{]}.

@@

Boden and McKendrick {[}12{]} propose a framework for ethical public health modelling based on the criteria of independence (concerning how modeller subjectivity shapes model design), beneficence / non-maleficence (concerning model quality and utility), transparency (concerning the need for policymakers to reliably evaluate model strengths and weaknesses) and justice (concerning the social obligations of modellers to consider and communicate ethical issues about model use). However, these criteria are not specific to the computational model, but are designed to apply to a modelling project as a whole, including a model's conceptual and mathematical representations. The authors propose 13 questions to evaluate ethical risk across the four criteria, but of these, only ``is the model code open source or available on request'' (for transparency) and ``has the model been verified, i.e., does it do what the modeller wants it to do?'' (for beneficence / non-maleficence) are specific to model's representation as software.

To identify ethical issues that are specific to CHEMs, we first considered the distinct phases in a software development project. Software development lifecycle (SDLC) models take different approaches to the inclusion, naming, definition, sequencing and iteration of project phases {[}14{]}. However, SDLC models typically have components that can principally map to the concepts of \emph{planning} (e.g., identification of system / user requirements and design specifications), \emph{implementation} (e.g., development and testing) and \emph{release} (e.g., deployment, system integration, maintenance and support). For simplicity, we propose one overarching ethical responsibility for CHEM project teams for each phase. These responsibilities are to:

@@

Misalignment between the values of computational model developers and those of the population groups affected by decisions based on their models presents significant ethical risks {[}15,16{]}. The value judgments of health economic modellers are rarely adequately specified, omissions that may lead to socially unacceptable policy recommendations {[}16{]}. These value judgments influence the assumptions, selection of model features and standards for evidence that shape health economic model projects {[}17{]}. For example, as illustrated in Table \ref{tab:proscons}, project resource constraints may require model developers to make trade-offs when agreeing CHEM \textbf{user requirements and design specifications}. These trade-offs in turn determine how straightforward it is for a CHEM to be subsequently amended to reflect alternative value judgments relating to model conceptualisation and mathematical formalisation (Table \ref{tab:timelygls}).

Modellers have duties both to ensure a computational model is \textbf{fit for purpose} and to provide potential third-party users with the means of assessing its adequacy for their proposed uses {[}6,12,18,19{]}. However, it is common for health economic models to have serious methodological flaws {[}20,21{]}; insufficient validation {[}22--24{]}, poor reproducibility {[}25--27{]}; and undeclared errors {[}28{]}. Appropriate computational implementation choices can help address many of these shortcomings, for example by automating quality assurance checks and facilitating manual reviews by third parties.

A CHEM will have limited \textbf{social benefit} if it is rarely used, if mis-used or when its acceptability and adequacy rapidly decay. Reuse of CHEMs as components of other models can potentially make model development more efficient {[}29,30{]}. However, health economic models face challenges related to transferability across jurisdictions {[}30{]} that create barriers to reuse. Without ongoing maintenance, a CHEM risks becoming less reliable with time {[}30{]} and is at risk of being deployed for purposes for which it is poorly suited {[}13{]}. Currently, health economic models are rarely implemented computationally in a manner that facilitates routine updates {[}31{]}, thus limiting the temporal window within which a CHEM can be validly applied.

@@

\global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}

\global\setlength{\Oldtabcolsep}{\tabcolsep}

\setlength{\tabcolsep}{0pt}

\renewcommand*{\arraystretch}{1.5}

\providecommand{\ascline}[3]{\noalign{\global\arrayrulewidth #1}\arrayrulecolor[HTML]{#2}\cline{#3}}

\begin{longtable}[c]{|p{1.20in}|p{1.60in}|p{1.60in}|p{1.60in}}

\caption{Transparent,\ reusable\ and\ updatable\ (TRU)\ computational\ health\ economic\ models\ (CHEMs)\ are\ suggestive\ of\ ethical\ modelling\ practice}\label{tab:timelygls}\\

\ascline{1pt}{666666}{1-4}

\multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\centering}m{\dimexpr 1.2in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textbf{\ }}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 1pt}>{\centering}m{\dimexpr 1.6in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textbf{Social\ acceptability}}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 1pt}>{\centering}m{\dimexpr 1.6in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textbf{Fitness\ for\ purpose}}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 1pt}>{\centering}m{\dimexpr 1.6in+0\tabcolsep}!{\color[HTML]{666666}\vrule width 1pt}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textbf{Beneficial\ use}}}} \\

\ascline{1pt}{666666}{1-4}\endfirsthead \caption[]{Transparent,\ reusable\ and\ updatable\ (TRU)\ computational\ health\ economic\ models\ (CHEMs)\ are\ suggestive\ of\ ethical\ modelling\ practice}\label{tab:timelygls}\\

\ascline{1pt}{666666}{1-4}

\multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\centering}m{\dimexpr 1.2in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textbf{\ }}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 1pt}>{\centering}m{\dimexpr 1.6in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textbf{Social\ acceptability}}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 1pt}>{\centering}m{\dimexpr 1.6in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textbf{Fitness\ for\ purpose}}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 1pt}>{\centering}m{\dimexpr 1.6in+0\tabcolsep}!{\color[HTML]{666666}\vrule width 1pt}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textbf{Beneficial\ use}}}} \\

\ascline{1pt}{666666}{1-4}\endhead

>>>>

%Version 2.1 April 2023
% See section 11 of the User Manual for version history
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[sn-vancouver,Numbered,pdflatex]{sn-jnl}

%%%% Standard Packages
%%<additional latex packages if required can be included here>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published
%%%%  by Springer Nature. The guidance has been prepared in partnership with
%%%%  production teams to conform to Springer Nature technical requirements.
%%%%  Editorial and presentation requirements differ among journal portfolios and
%%%%  research disciplines. You may find sections in this template are irrelevant
%%%%  to your work and are empowered to omit any such section if allowed by the
%%%%  journal you intend to submit to. The submission guidelines and policies
%%%%  of the journal take precedence. A detailed User Manual is available in the
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

%% Per the spinger doc, new theorem styles can be included using built in style, 
%% but it seems the don't work so commented below
%\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

%% \theoremstyle{thmstyletwo}%
\theoremstyle{remark}
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

%% \theoremstyle{thmstylethree}%
\theoremstyle{definition}
\newtheorem{definition}{Definition}%

\usepackage{float}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}
\usepackage{lscape}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\usepackage[none]{hyphenat}
\usepackage[export]{adjustbox}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{array}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{booktabs} 
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage[normalem]{ulem}
\usepackage{threeparttablex}
\usepackage{makecell}

\raggedbottom

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

\begin{document}

\title[Transparent, reusable and updatable computational models]{A prototype software framework for transparent, reusable and updatable computational health economic models}

%%=============================================================%%
%% Prefix	-> \pfx{Dr}
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% NatureName	-> \tanm{Poet Laureate} -> Title after name
%% Degrees	-> \dgr{MSc, PhD}
%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate}
%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
%%=============================================================%%

\author*[1,2]{\pfx{Dr.} \fnm{Leading} \spfx{van} \sur{Author} \sfx{III} \tanm{Poet Laureate} \dgr{MSc, PhD}}\email{\href{mailto:abc@def}{\nolinkurl{abc@def}}}

\author[2]{\fnm{Second} \sur{Author} }

  \affil*[1]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\city{City}, \country{Country}, \postcode{100190}, \state{State}, \street{Street}}}
  \affil*[2]{\orgname{Other Organisation}}

\abstract{\textbf{Summary: } Most health economic analyses are undertaken with the aid of computers. However, the ethical dimensions of implementing health economic models as software (or computational health economic models (CHEMs)) are poorly understood. We propose that developers and funders of CHEMs share ethical responsibilities to (i) establish socially acceptable user requirements and design specifications; (ii) ensure fitness for purpose; and (iii) support socially beneficial use. We further propose that a transparent (T), reusable (R) and updatable (U) CHEM is suggestive of a project team that has largely fulfilled these responsibilities. We propose six criteria for assessing CHEMs: (T1) software files are open access; (T2) project team contributions and judgments are easily identified; (R1) programming practices promote generalisability and transferability; (R2) licenses restrict only unethical reuse; (U1) maintenance infrastructure is in place; and (U2) new releases are systematically retested and appropriately deprecated. To facilitate CHEMs that meet TRU criteria, we have developed a prototype software framework in the open-source programming language R. The framework comprises six code libraries for authoring CHEMs, supplying CHEMs with data and undertaking analyses with CHEMs. The prototype software framework integrates with services for software development and research data archiving. We determine that an initial set of youth mental health CHEMs we developed with the prototype software framework wholly meet criteria T1-2, R1-2 and U1 and partially meet criterion U2. Our assessment criteria and prototype software framework can help inform and improve ethical implementation of CHEMs. Resource barriers to ethical CHEM practice should be addressed by research funders. \newline \newline \textbf{Code: } Visit \url{https://www.ready4-dev.com} for more information about how to find, install and apply the prototype software framework and CHEMs developed with it. \newline \newline}

\keywords{key, dictionary, word}

\pacs[JEL Classification]{D8, H51}
\pacs[MSC Classification]{35A01, 65L10}

\maketitle

@@

Health economics is a discipline concerned with problems that arise due to scarce resources, such as how to value health and healthcare, allocate healthcare budgets and configure health services \citep{wagstaff2012four}. In seeking to solve these problems, health economists typically use models which are simplified and selective representations of systems that are believed to influence human health. A health economic model should be capable of being described using words and figures (a conceptual model), equations (a mathematical model) and software (a computational model). Health economic scientific manuscripts typically describe a model using its conceptual and mathematical representations but report results that have been generated by its computational representation (i.e., the format that allows computers to perform calculations). The conceptual, mathematical and computational representations of a model are assumed to be isomorphic. However, the ease of independently assessing the validity of this assumption depends in part on how a health economic model has been implemented computationally.

The set of software files implementing a health economic model's structure, data and algorithms can be called a computational health economic model (CHEM). CHEMs can be developed using commercial modelling software or open-source programming languages such as Python \citep{python2009} or R \citep{RCORE2022}. An advantage of commercial modelling software is that users often require limited or no programming skills to develop and apply CHEMs. However, open-source software programming languages may facilitate the development of CHEMs that are more transparent, reusable and updatable \citep{incerti2019r, Pouwels2022}. The decision about what type of software development platform to use must balance ethical considerations and resource constraints (see Table \ref{tab:proscons}).

Open-source CHEMs remain rare \citep{Feenstra2022, Emerson2019}. The software used to develop CHEMs was identified as a barrier to open-source model implementations by participants in a 2020 survey of health economists \citep{Pouwels2022}. Other reported barriers relate to difficulties in generalising models, updating models and transferring model data, concerns about the level of public access and lack of interest from decision-makers, and legal, confidentiality and security considerations \citep{Pouwels2022}.

Better software frameworks could potentially help address some of the barriers to open-source CHEM development. A software framework is a shared common technology used by developers to collaboratively author software and which is not typically visible to software end-users \citep{myllarniemi2018development}. A software framework provides a foundation for developing multiple software applications with shared resources (e.g., code and data files), that can be modified to suit specific needs. Advantages of using software frameworks include facilitating code reuse and extension, promoting good programming practice and the capability to provide enhanced functionality and performance without additional effort by developers \citep{edwin2014software}. Software frameworks have widely been developed and implemented in data science, for example, PyTorch \citep{PyTorch2019} for machine learning using Python.

A high-level software framework for implementing open-source CHEMs in R has previously been developed with the primary aim of improving model transparency \citep{alarid2019need}. However, to gain widespread adoption, a software framework for CHEMs may need to provide tools to help model developers meet a broader range of ethical practice requirements. There are multiple challenges to overcome. Firstly, the ethical responsibilities of public health computational modellers remain poorly understood \citep{10.3389/fpubh.2017.00068} and the computational modelling landscape is changing fast \citep{calder2018computational}. Secondly, software frameworks can be challenging and time consuming to create and may become excessively complex over time \citep{edwin2014software}. Thirdly, software frameworks can be difficult to learn, often requiring model developers to undergo specialist training \citep{myllarniemi2018development}.

@@

Boden and McKendrick \citep{10.3389/fpubh.2017.00068} propose a framework for ethical public health modelling based on the criteria of independence (concerning how modeller subjectivity shapes model design), beneficence / non-maleficence (concerning model quality and utility), transparency (concerning the need for policymakers to reliably evaluate model strengths and weaknesses) and justice (concerning the social obligations of modellers to consider and communicate ethical issues about model use). However, these criteria are not specific to the computational model, but are designed to apply to a modelling project as a whole, including a model's conceptual and mathematical representations. The authors propose 13 questions to evaluate ethical risk across the four criteria, but of these, only ``is the model code open source or available on request'' (for transparency) and ``has the model been verified, i.e., does it do what the modeller wants it to do?'' (for beneficence / non-maleficence) are specific to model's representation as software.

To identify ethical issues that are specific to CHEMs, we first considered the distinct phases in a software development project. Software development lifecycle (SDLC) models take different approaches to the inclusion, naming, definition, sequencing and iteration of project phases \citep{ruparelia2010software}. However, SDLC models typically have components that can principally map to the concepts of \emph{planning} (e.g., identification of system / user requirements and design specifications), \emph{implementation} (e.g., development and testing) and \emph{release} (e.g., deployment, system integration, maintenance and support). For simplicity, we propose one overarching ethical responsibility for CHEM project teams for each phase. These responsibilities are to:

@@

Misalignment between the values of computational model developers and those of the population groups affected by decisions based on their models presents significant ethical risks \citep{thompson2022escape, duckett2022journey}. The value judgments of health economic modellers are rarely adequately specified, omissions that may lead to socially unacceptable policy recommendations \citep{duckett2022journey}. These value judgments influence the assumptions, selection of model features and standards for evidence that shape health economic model projects \citep{HARVARD2020112975}. For example, as illustrated in Table \ref{tab:proscons}, project resource constraints may require model developers to make trade-offs when agreeing CHEM \textbf{user requirements and design specifications}. These trade-offs in turn determine how straightforward it is for a CHEM to be subsequently amended to reflect alternative value judgments relating to model conceptualisation and mathematical formalisation (Table \ref{tab:timelygls}).

Modellers have duties both to ensure a computational model is \textbf{fit for purpose} and to provide potential third-party users with the means of assessing its adequacy for their proposed uses \citep{Erdemir2020, Feenstra2022, thompson2019escape, 10.3389/fpubh.2017.00068}. However, it is common for health economic models to have serious methodological flaws \citep{carletto_zanuzzi_sammarco_russo_2020, WONDER2015467}; insufficient validation \citep{Ghabri2019, kolovos2017model, haji2013model}, poor reproducibility \citep{Jalali2021, McManus2019, Bermejo2017}; and undeclared errors \citep{Radeva2020}. Appropriate computational implementation choices can help address many of these shortcomings, for example by automating quality assurance checks and facilitating manual reviews by third parties.

A CHEM will have limited \textbf{social benefit} if it is rarely used, if mis-used or when its acceptability and adequacy rapidly decay. Reuse of CHEMs as components of other models can potentially make model development more efficient \citep{Arnold2010, garcia2021cost}. However, health economic models face challenges related to transferability across jurisdictions \citep{garcia2021cost} that create barriers to reuse. Without ongoing maintenance, a CHEM risks becoming less reliable with time \citep{garcia2021cost} and is at risk of being deployed for purposes for which it is poorly suited \citep{calder2018computational}. Currently, health economic models are rarely implemented computationally in a manner that facilitates routine updates \citep{Sampson2017}, thus limiting the temporal window within which a CHEM can be validly applied.

@@

Taking steps to make computational models \textbf{transparent} is an important part of ethical public health modelling practice \citep{10.3389/fpubh.2017.00068}. Guidance on transparency in health economic modelling recommend that model code and data should be clearly documented \citep{Eddy2012}. Notably, the same guidelines, published over ten years ago, do not recommend sharing model code and data. However, more recent healthcare modelling guidance does recommend public dissemination of such artefacts \citep{Erdemir2020}. Online repository services such as Zenodo \citep{Zenodo2013} and Dataverse \citep{Dataverse2007} provide persistent storage solutions that generate a Digital Object Identifier (DOI) for each code and data collection. Ensuring that calculations are correct and consistent with model specifications is an essential part of CHEM quality assurance \citep{techver2019}. The extensiveness of such verification checks can be reported using the concept of code coverage \citep{ERICWONG2010188} - the proportion of model code that has been explicitly tested. Tests should ideally combine both unit tests (to verify that small, isolated sections of code produce the correct output when run independently) and acceptance tests (to verify that the correct output is produced when multiple code components are run together to perform tasks that meet core user-requirements \citep{martin2003agile}). The nature and extent of individual model authorship contributions can become unclear when models are implemented over longer time-frames with a large and changing group of collaborators \citep{thompson2022escape}. This issue can be addressed by use of online repository services such as GitHub \citep{github2007}, that provide citation tools and can transparently record all individual code contributions to a modelling project over its lifecycle.

A CHEM that is \textbf{reusable} also signals ethical modelling practice. Although making a CHEM's code, data and documentation publicly available is increasingly considered good practice, it is insufficient to enable model re-use. Key concerns for health economists when considering whether to reuse a model are generalisability (application without adaptation) and transferability (selective reuse and/or modification of model components) \citep{RN39}. Writing model algorithms as collections of functions (short, self-contained and reusable software routines that each perform a discrete task) is good scientific computing practice \citep{Wilson_2017} and promotes selective reuse. Computational implementations that store model code and data in distinct files and locations (as opposed to embedding data such as parameter values into source code) are easier to selectively modify. Modular implementations enable models to be constructed from multiple independently reusable and replaceable sub-models (modules) \citep{pan2021modular}. The programming concept inheritance \citep{8717448}, deployed in object-oriented programming approaches, can facilitate the duplication and selective modification of models. Granting permissions to others to use, test and adapt models and their components, can be facilitated by two broad categories of open source licenses. Some guidance strongly recommends the use of permissive licensing \citep{Wilson_2017} that provides users with great flexibility as to the purposes (including commercial) for which content can be re-used. Alternatively, copyleft licenses \citep{copyleft2022} can be used to require that derivative works created by content users remain open-source.

Models should be \textbf{updatable} so that they remain valid for longer, evolving as new evidence emerges and the systems being modelled change \citep{Jenkins2021, garcia2021cost}. Ensuring that a model is regularly reviewed to identify and implement required improvements is a recommended defence against model validity decay \citep{calder2018computational}. Sustainable maintenance of open source research software requires both a core development team and an active user community \citep{info:doi/10.2196/20028}. Online communities can be an efficient means of engaging model users in testing each version of a model, identifying issues and suggesting improvements. Services such as GitHub \citep{github2007} provide collaborative code development tools \citep{MERGEL2015464} that help elicit, integrate and reconcile contributions from multiple contributors and to ensure each update is uniquely identifiable and retrievable. It is also important that verification checks are rerun with each model update, a task that can be automated using the software development practice of continuous integration \citep{CI2017}. The risk of software revisions having unintended consequences for third party users can be mitigated through the use of deprecation conventions \citep{zhou2016api} that take an informative and staged approach to retiring outdated CHEMs.
